---
output:
  github_document:
    toc: true
    toc_depth: 2
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# jevr

<!-- badges: start -->
[![R-CMD-check](https://github.com/simxnherrera/jevr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/simxnherrera/jevr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

jevr is a small native R client for [JEV / TypeSafe System One](https://docs.typesafe.ai/concepts/system-one).
It turns a state and a set of typed questions into structured decisions:

```text
state + questions -> model -> typed answers
```

System One is a decision interface, not a conversational chat interface. The
package therefore focuses on typed questions, batched requests, probabilities,
confidence, and provider-independent response objects. It does not expose chat
history, token streaming, or tool calling.

## Installation

Install the development version from GitHub:

```{r, eval=FALSE}
install.packages("pak")
pak::pak("simxnherrera/jevr")
```

For local development:

```{r, eval=FALSE}
pak::pak("devtools")
devtools::load_all()
```

## Credentials

Use an environment variable for the provider you select. Do not put API keys
in package code, tests, or committed files.

```{r, eval=FALSE}
usethis::edit_r_environ()

# For direct TypeSafe requests:
TYPESAFE_API_KEY=your_typesafe_key

# For OpenRouter requests:
OPENROUTER_API_KEY=your_openrouter_key
```

The key is read when jev_ask() is called. The package never prints it in
errors or stores it in the repository.

## First request

The public API has four functions: jev_ask(), jev_choice(), jev_score(), and
jev_noul().

```{r, eval=FALSE}
library(jevr)

questions <- list(
  department = jev_choice(
    instructions = "Which team should handle this?",
    criteria = c(
      billing = "Payments, invoicing, refunds",
      technical = "Bugs, outages, integrations",
      sales = "Pricing, upgrades, new accounts"
    )
  ),
  urgent = jev_noul(
    instructions = "Does this request convey urgency?"
  )
)

result <- jev_ask(
  state = "Help! My payouts have been failing for three days.",
  questions = questions
)

result$model
result$answers$department$choice
result$answers$department$probabilities
result$answers$urgent$noul
```

Question IDs come from the names of the questions list. They are preserved
in the request and in result$answers, so multiple questions can be sent in a
single request without losing the mapping between a question and its answer.

## Typed questions

### Choice

jev_choice() selects one named option and preserves the full probability
distribution and confidence:

```{r, eval=FALSE}
question <- jev_choice(
  instructions = "Which priority applies?",
  criteria = c(
    low = "Can wait",
    high = "Needs attention soon"
  )
)
```

### Score

jev_score() returns an ordered score, its legend, probabilities, and
confidence:

```{r, eval=FALSE}
question <- jev_score(
  instructions = "How severe is the issue?",
  criteria = c(
    "Minor inconvenience",
    "Material disruption",
    "Service unavailable"
  )
)
```

### Noul

jev_noul() returns the continuous value between 0 and 1. It does not silently
convert that value to TRUE or FALSE:

```{r, eval=FALSE}
question <- jev_noul(
  instructions = "Does the request require immediate action?",
  criteria = list(
    true = "The request describes an immediate operational risk",
    false = "The request can wait without immediate harm"
  )
)
```

## Providers

### TypeSafe

TypeSafe is the default provider. jevr sends a batched request to the
official System One endpoint and uses the stable jev-latest model alias by
default.

```{r, eval=FALSE}
result <- jev_ask(
  state = list(
    customer = "Acme",
    message = "My payouts have been failing for three days."
  ),
  questions = questions,
  provider = "typesafe"
)
```

state may be a character value, a named list representing a JSON object, or an
unnamed list representing a JSON array. The package keeps those structures as
structured JSON when the provider supports them.

### OpenRouter

OpenRouter is available through its Decisions endpoint. Select it without
changing the question definitions:

```{r, eval=FALSE}
result <- jev_ask(
  state = "Help! My payouts have been failing for three days.",
  questions = questions,
  provider = "openrouter"
)
```

OpenRouter uses ~typesafe/jev-latest by default and authenticates with
OPENROUTER_API_KEY. Its current Decisions schema accepts textual instructions
and criteria; use the direct TypeSafe provider when you need structured
instruction or criteria values.

## Development

The package follows the workflow from [R Packages (2e)](https://r-pkgs.org/):

```{r, eval=FALSE}
devtools::load_all()
devtools::document()
devtools::test()
devtools::check()
```

Normal tests mock the HTTP layer and do not require credentials or make
network requests. Paid live integration tests are skipped on CRAN and unless
you explicitly set JEVR_RUN_INTEGRATION_TESTS=true together with the relevant
provider key.

## Status

jevr is an early development version. The initial API and provider contracts
are intentionally small; the package does not yet provide a chat interface,
streaming, tool calling, or bulk classification helpers.
