---
output:
  github_document:
    toc: true
    toc_depth: 2
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# jevr

<!-- badges: start -->
[![R-CMD-check](https://github.com/simxnherrera/jevr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/simxnherrera/jevr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`jevr` is an R client for [JEV / TypeSafe System One](https://docs.typesafe.ai/concepts/system-one).
It sends a state and a set of typed questions to a model and returns structured
answers that can be used directly in R.

JEV is designed for structured decisions rather than conversational chat. The
package supports multiple questions in one evaluation, probabilities,
confidence, usage, and provider metadata.

## Installation

Install the development version from GitHub:

```{r, eval=FALSE}
install.packages("pak")
pak::pak("simxnherrera/jevr")
```

Once `jevr` is available on CRAN, the stable version can be installed with:

```{r, eval=FALSE}
install.packages("jevr")
```

## Authentication

Set the API key for the provider you want to use in your `.Renviron` file:

```{r, eval=FALSE}
usethis::edit_r_environ()
```

```text
# Direct TypeSafe requests
TYPESAFE_API_KEY=your_typesafe_key

# OpenRouter requests
OPENROUTER_API_KEY=your_openrouter_key
```

Keys are read when `jev_ask()` is called and are not included in returned
objects.

## Basic usage

Define one or more questions, then pass them together with the state to
`jev_ask()`:

```{r, eval=FALSE}
library(jevr)

questions <- list(
  department = jev_choice(
    instructions = "Which team should handle this?",
    criteria = c(
      billing = "Payments, invoicing, refunds",
      technical = "Bugs, outages, integrations",
      sales = "Pricing, upgrades, new accounts"
    )
  ),
  urgent = jev_noul(
    instructions = "Does this request convey urgency?"
  )
)

result <- jev_ask(
  state = "Help! My payouts have been failing for three days.",
  questions = questions
)

result$answers$department$choice
result$answers$department$probabilities
result$answers$urgent$noul
```

The names in `questions` become the IDs in `result$answers`. This makes it
possible to send several questions in one request without losing the mapping
between each question and its answer.

`jev_ask(state, questions)` represents one evaluation of one state. For a
collection of states, use `jev_map(states, questions)`: each state is evaluated
independently, with bounded HTTP concurrency. `jev_map()` is not model-level
batching of multiple records into one state or one request.

`state` can be a character string, a named list representing a JSON object, or
an unnamed list representing a JSON array.

## Question types

### Choice

`jev_choice()` selects one option from a named set. The answer contains the
selected option, the probability for each option, and model confidence.

```{r, eval=FALSE}
jev_choice(
  instructions = "Which priority applies?",
  criteria = c(
    low = "Can wait",
    high = "Needs attention soon"
  )
)
```

### Score

`jev_score()` places the state along an ordered scale of two to ten levels. The
answer contains the score, its legend, probabilities, and confidence.

```{r, eval=FALSE}
jev_score(
  instructions = "How severe is the issue?",
  criteria = c(
    "Minor inconvenience",
    "Material disruption",
    "Service unavailable"
  )
)
```

### Noul

`jev_noul()` estimates how strongly a statement is true and returns a numeric
value between 0 and 1. The value is not converted to `TRUE` or `FALSE`.

```{r, eval=FALSE}
jev_noul(
  instructions = "Does the request require immediate action?",
  criteria = list(
    true = "The request describes an immediate operational risk",
    false = "The request can wait without immediate harm"
  )
)
```

## Providers

### TypeSafe

TypeSafe is the default provider. It uses the `jev-latest` model alias unless a
different `model` is supplied:

```{r, eval=FALSE}
result <- jev_ask(
  state = list(
    customer = "Acme",
    message = "My payouts have been failing for three days."
  ),
  questions = questions,
  provider = "typesafe"
)
```

The direct TypeSafe provider supports structured JSON values for `state`,
`instructions`, and criteria.

### OpenRouter

OpenRouter can be selected without changing the question definitions:

```{r, eval=FALSE}
result <- jev_ask(
  state = "Help! My payouts have been failing for three days.",
  questions = questions,
  provider = "openrouter"
)
```

It uses `~typesafe/jev-latest` by default and authenticates with
`OPENROUTER_API_KEY`. Structured JSON instructions and criteria are preserved
when the provider contract accepts them.

### ellmer and ellmercodex (optional)

The same questions and `jev_spec()` objects can be evaluated with an ellmer
`Chat`. Supply a factory that creates a new chat for each evaluation and an
explicit, data-only description of its semantic configuration:

```{r, eval=FALSE}
backend <- jev_llm(
  chat = function() {
    ellmercodex::chat_codex(model = "gpt-6-luna", effort = "medium")
  },
  model = "gpt-6-luna",
  inference_options = list(reasoning_effort = "medium")
)

result <- jev_ask(
  state = "Help! My payouts have been failing for three days.",
  questions = questions,
  backend = backend
)

results <- jev_map(
  states = c(first = "Payouts fail", second = "A refund is pending"),
  questions = questions,
  backend = backend,
  concurrency = 1,
  on_result = function(result) {
    # Persist the terminal result in caller-owned storage.
  }
)
```

Creating `backend` does not create a chat or contact the service. ellmer and
ellmercodex are optional packages; ellmercodex owns authentication and
transport. Authenticate separately using its documented workflow. Without
`backend`, TypeSafe and OpenRouter continue to use their native Decisions APIs.

All questions in one evaluation share a single structured call, with their
IDs, order, JSON instructions and criteria preserved. The prompt treats the
state as source data and instructs the model to disregard instructions in that
material. Each question must be answered independently. This prompt instruction
is not a guarantee of resistance to adversarial source material.

Choice and Score return distributions; Noul returns probability of true.
jevr selects the Choice option with highest probability, breaking ties by the
original criteria order. Score is the weighted mean over levels `0` through
`k - 1`: `sum(level * probability) / sum(probability)`. Distributions must contain
exactly the requested keys, finite numbers in `[0, 1]`, and sum to one within
`probability_tolerance` (default `1e-6`, maximum `1e-3`). Accepted rounded
probabilities are preserved; invalid distributions are rejected without
renormalization. The Score denominator accounts only for accepted rounding.

These are self-reported LLM probabilities. Each answer and response metadata
records `probability_source = "llm_self_report"` and
`confidence_status = "absent"`; Choice and Score have explicit
`confidence = NULL`. Native responses still require confidence. `result$raw`
is the structured object returned by ellmer; `result$answers` contains the
validated jevr representation, including locally calculated decisions.

Metadata records the upstream provider name from public
`Chat$get_provider()@name`, requested model, model from public
`Chat$get_model()` when available, adapter version, inference options and
`configuration`. The model getter describes ellmer's model; it does not certify
an upstream resolved model ID. Public `Chat$get_tokens()` supplies available
usage and cost; unavailable usage is `NA`, and unavailable cost is omitted.
For ellmercodex, `metadata$upstream_provider` is `"codex"`. Only the provider
name is retained; its object and credentials are not stored. Unavailable names
remain `NULL`. The name participates in the execution identity after chat
construction. States that never construct a chat (invalid, not admitted, or
factory failures) have `execution_id = NA`. No missing metadata is inferred
from the requested configuration.

The declared settings are not applied automatically to the factory. Describe
additional semantic settings (for example, `system_prompt` or `api_args`) in
`configuration`, and keep the description consistent with the factory.
Changing these settings changes the execution identity. Factories, Chat objects,
environments and credentials are never included in provenance or hashes.

This route supports only `concurrency = 1`. Explicit `provider`, `model`,
`timeout`, `retry_budget` (map only), or positive `max_retries` conflict with the
backend and are rejected. Omitted native defaults do not control ellmer.
Timeouts and retries belong to the transport: `chat_codex()` currently provides
no public per-evaluation timeout override, so jevr cannot promise one.
`rate_limit` spaces sequential admissions; ellmer parallel/batch helpers are
not used. The existing callbacks, partial results and result-set operations
remain available.

The ledger records one jevr invocation per admitted evaluation with elapsed
duration and timestamps. HTTP status and the transport's internal attempt count
are unknown. `summary(results)$llm_invocations` counts these invocations;
`http_attempts` counts only observed native HTTP attempts. Transport failures
retain the original condition as the cause in `jev_ask()` and a data-only cause
record in map errors.

Tests run without network by default. The optional real integration test
requires `JEVR_RUN_ELLMERCODEX_INTEGRATION=true` and a separately authenticated
ellmercodex session. Enabling it uses the subscription and sends one evaluation
with Choice, Score and Noul. It does not perform authentication for you.

## Versioned specifications and multiple states

Use `jev_spec()` when a definition must be identified and compared over time:

```{r, eval=FALSE}
spec <- jev_spec(
  name = "claim_validation",
  version = "1.0.0",
  questions = questions
)
```

The spec contains only data and a recalculable SHA-256 hash. It does not store
credentials, provider clients, callbacks, or operational settings.

Use `jev_definition_manifest()` to inspect the canonical data-only definition.
`jev_spec_hash()` and `jev_state_hash()` provide stable hashes for a definition
and a state. `jev_execution_id()` combines the state, definition, provider,
model, and inference semantics into a reproducible execution identity:

```{r, eval=FALSE}
jev_definition_manifest(spec)
jev_spec_hash(spec)
jev_state_hash(list(document = "First document"))
jev_execution_id(
  state = list(document = "First document"),
  definition = spec,
  provider = "typesafe",
  model = "jev-latest"
)
```

`jev_map()` evaluates explicit states independently with bounded concurrency
and returns a named `jev_result_set`. Failures stay attached to their state
instead of aborting the whole collection:

```{r, eval=FALSE}
states <- c(
  document_001 = "First document",
  document_002 = "Second document"
)

results <- jev_map(
  states,
  spec,
  concurrency = 4,
  on_result = function(result) {
    # Persist the terminal result in caller-owned storage.
  }
)

as.data.frame(results)
attr(results, "requests")
```

The callback is the persistence boundary: jevr does not manage a database,
checkpoint files, or a distributed job queue. For large sources, read chunks
outside the package and call `jev_map()` for each chunk. `concurrency` limits
requests in flight; `rate_limit` controls admission separately.

When the provider returns a valid response envelope but an individual answer
cannot be validated, `jev_map()` returns `status = "partial"`, keeps the valid
answers, and records the invalid questions in `question_errors`. The raw
response remains available for audit. `attr(results, "requests")` contains one
row per physical native HTTP attempt; `duration_seconds` is the per-response HTTP
timing reported by httr2 and can be `NA` when the backend does not provide a
reliable value. Per-request ledger timestamps are `NA` because the public
parallel API does not expose reliable start and finish timestamps after the
wave returns.

The development cancellation probe also documents a limitation of the current
public httr2 API: when an active wave is interrupted and httr2 drains every
active request, it can return a complete response list without an explicit
interrupted flag. jevr handles observable `NULL` results conservatively, but
cannot distinguish that drained wave from a normal wave. Closing that gate
requires a transport adapter that returns an explicit interrupted flag; jevr
does not introduce a second transport backend in this release.

## Response object

`jev_ask()` returns an object of class `jev_response` with these main fields:

- `model`: reported model, or the requested model when ellmer metadata is unavailable.
- `answers`: named answers corresponding to the supplied questions.
- `usage`: input and output token counts returned by the provider.
- `metadata`: provider and upstream request metadata.
- `raw`: the unmodified provider response.

Use `?jev_ask`, `?jev_llm`, `?jev_map`, `?jev_spec`, `?jev_definition_manifest`,
`?jev_spec_hash`, `?jev_state_hash`, `?jev_execution_id`, `?jev_choice`,
`?jev_score`, and `?jev_noul` for the complete argument reference.
