---
title: "Composite scoring"
---

Sometimes a single number must summarize many qualities: lead quality, risk,
candidate fit, content health. Asking one question for the overall score
hides your policy inside a prompt. The composite pattern from the TypeSafe
docs does the opposite: ask many small questions that each measure one thing,
and combine them with weights in your own code. This article shows that
workflow with jevr. Chunks are not evaluated here.

```{r, eval = FALSE}
library(jevr)
```

## Atomic Score questions

Each question should measure one attribute, with levels that read from low to
high and describe observable differences. Four to five levels is plenty.

```{r, eval = FALSE}
questions <- list(
  budget = jev_score(
    "How clearly does the lead indicate a budget?",
    c(
      "No mention of budget",
      "Vague hints about cost sensitivity",
      "Gives a rough range",
      "States a concrete, approved budget"
    )
  ),
  timeline = jev_score(
    "How soon does the lead intend to buy?",
    c(
      "No timeline",
      "Exploring, more than a year out",
      "Planning within the year",
      "Needs a solution within a month"
    )
  ),
  fit = jev_score(
    "How well does the request match what we sell?",
    c(
      "Unrelated to our product",
      "Adjacent; would need custom work",
      "Mostly a match",
      "Exactly our core use case"
    )
  )
)
```

The questions are independent: they are evaluated together against the same
state, but one answer does not influence another.

## Run once

```{r, eval = FALSE}
leads <- c(
  lead_a = "We have 40k approved for a CRM and want to launch next month.",
  lead_b = "Just curious how this compares to spreadsheets, no rush.",
  lead_c = "Can you build us a custom drone fleet manager?"
)

results <- jev_map(leads, questions)
```

## Collect the components

In long format a Score question's `score` is the probability-weighted
expectation, repeated on each level row. Take it once per state and question
and arrange it as a matrix.

```{r, eval = FALSE}
long <- as.data.frame(results, format = "long")
long <- long[long$status == "success", ]

components <- tapply(long$score, list(long$state_id, long$question_id), mean)
components
```

The values are on the question's own scale (here 1 to 4, possibly
fractional). Rescale each to 0 to 1 so weights are comparable:

```{r, eval = FALSE}
n_levels <- c(budget = 4, timeline = 4, fit = 4)
unit <- sweep(components - 1, 2, (n_levels - 1)[colnames(components)], "/")
```

## Weights live in R

```{r, eval = FALSE}
composite <- function(unit, weights) {
  weights <- weights[colnames(unit)] / sum(weights[colnames(unit)])
  drop(unit %*% weights)
}

weights <- c(budget = 0.5, timeline = 0.2, fit = 0.3)
priority <- composite(unit, weights)
sort(priority, decreasing = TRUE)
```

Because the model only produced the components, changing your mind costs
nothing. A sales-led team and a product-led team can rank the same leads
differently, and you can try new weights as often as you like without another
request:

```{r, eval = FALSE}
sales_led   <- c(budget = 0.6, timeline = 0.3, fit = 0.1)
product_led <- c(budget = 0.1, timeline = 0.2, fit = 0.7)

data.frame(
  sales   = composite(unit, sales_led),
  product = composite(unit, product_led)
)
```

You can also add rules the model should never decide: a hard floor on `fit`,
a bonus for a known account, a cap for leads that already churned.

```{r, eval = FALSE}
priority[unit[, "fit"] < 0.2] <- 0
```

## Keep the evidence

Store the components next to the composite so a reviewer can see why a lead
scored the way it did, and which model version produced the numbers.

```{r, eval = FALSE}
report <- data.frame(
  lead = rownames(unit),
  unit,
  priority = priority,
  model = unique(long$model)[1],
  row.names = NULL
)
```

If a run mixes model versions (`jev_model_drift_warning`), the components are
not strictly comparable; re-run pinned to one versioned model. Failed states
are absent from `components` because of the status filter, so compare
`rownames(components)` with `names(results)` to find them.

## Rules of thumb

- One attribute per question; split any question that contains "and".
- Anchor every level with something observable, not an adjective.
- Use the same number of levels where you want to compare questions
  directly, and rescale when you do not.
- Keep weights, thresholds and tie-breaking in code under version control.
