---
title: "Consistency and uncertainty"
---

A System One answer is a probability distribution, not just a label, and the
distribution can tell you when not to trust the label. This article covers
three tools: checking whether a Choice answer depends on the *order* of its
options, turning confidence into routing bands, and the preflight warnings
that catch oversized requests. Chunks are not evaluated here.

```{r, eval = FALSE}
library(jevr)
```

## First-option bias

The Jev 1.13 notes on model jaggedness document a mild preference for the
first option of a Choice. For most questions it does not matter, but for
borderline ones it can flip the answer. The check is cheap: ask the same
Choice several times, with the options in different orders, in one request.

```{r, eval = FALSE}
questions <- list(
  team = jev_choice(
    "Which team should handle this request?",
    c(
      billing   = "Payments, invoices, refunds",
      technical = "Bugs, outages, integrations",
      sales     = "Pricing, upgrades, new accounts"
    )
  ),
  urgent = jev_noul("Does this request convey urgency?")
)

permuted <- jev_permute_choices(questions, n = 3, seed = 42)
names(permuted)
```

Every Choice is replaced by `n` copies with reordered options and ids such as
`team__perm1`; Score and Noul questions pass through. With a `seed` the orders
are reproducible, and when `n` does not exceed the number of options each
copy starts with a different option. Keep the returned object: it carries the
mapping back to the original ids, and a result only stores the expanded
definition.

A permuted spec is a different definition from the original, so it has a
different `jev_spec_hash()`; do not mix its results with unpermuted ones.

## Send and collapse

```{r, eval = FALSE}
response <- jev_ask(
  "The checkout page crashes whenever I pay by card.",
  permuted
)

jev_consistency(response, permuted)
```

`jev_consistency()` returns one row per state and original Choice question:

- `choice`: option with the highest mean probability across the copies;
- `agreement`: share of copies whose own choice equals `choice`
  (1 means order made no difference);
- `confidence` and `probabilities`: means over the answered copies;
- `choices` and `orders`: per-copy choice and the order each copy used;
- `status`: `"success"`, `"partial"` (some copies failed) or `"error"`.

It works the same on `jev_map()` results, with the same `permuted` object:

```{r, eval = FALSE}
tickets <- c(
  t1 = "The checkout page crashes whenever I pay by card.",
  t2 = "Do you offer a discount for annual billing?"
)
results <- jev_map(tickets, permuted)

check <- jev_consistency(results, permuted)
check[, c("state_id", "question_id", "choice", "agreement")]
unstable <- check[check$agreement < 1, ]
```

Unstable rows are candidates for human review, for a stronger model, or for a
rewrite of the option descriptions.

## Confidence bands

Confidence says how concentrated the distribution is, not whether the answer
is correct. Turn it into labels with `jev_band()`. It never picks thresholds
for you: `breaks` is required, and good values come from comparing against
your own labelled data.

```{r, eval = FALSE}
check$route <- jev_band(
  check$confidence,
  breaks = c(0, 0.5, 0.8, 1),
  labels = c("escalate", "review", "auto")
)
table(check$route, useNA = "ifany")
```

`jev_band()` also reads result objects directly, using `question` to name the
answer; values are banded with `cut()` semantics, so `right = TRUE` flips the
interval ends and the outermost end points are always included.

```{r, eval = FALSE}
names(permuted)  # the copy ids to band
jev_band(results, question = "team__perm1",
         breaks = c(0, 0.5, 0.8, 1), labels = c("escalate", "review", "auto"))
```

For a Noul the uncertain region is the middle, so use symmetric cuts:

```{r, eval = FALSE}
jev_band(
  results,
  question = "urgent",
  breaks = c(0, 0.3, 0.7, 1),
  labels = c("no", "uncertain", "yes")
)
```

The result is a factor (all levels kept, even when empty) named by state id.
It is `NA` for failed states, missing or invalid answers, answers with no
confidence (for example the unavailable-confidence sentinel from a gateway
fallback), and values outside the outer breaks. Decide explicitly what `NA`
should do in your pipeline; routing it to review is usually right.

## Preflight warnings

System One documents a 64k-token limit per request and a 32k-token limit for
the state plus the longest question. Choice questions allow at most 255
options and Score questions 2 to 10 levels; breaking those is an immediate
`jev_input_error`.

The size limits are approximate on the client. Before sending, jevr
estimates tokens as serialized characters divided by four and, if a limit
looks exceeded, signals a warning of class `jev_preflight_warning`. The
request is still sent, because the estimate can be wrong in either direction.
`jev_map()` collects affected states into one warning listing their ids.

```{r, eval = FALSE}
withCallingHandlers(
  jev_map(very_long_documents, questions),
  jev_preflight_warning = function(w) {
    message("Large request: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
```

Permuting multiplies question size, so check large Choice sets after
`jev_permute_choices()`. If the estimate is a nuisance in a pipeline that you
have already validated, switch it off for the session:

```{r, eval = FALSE}
options(jevr.preflight = FALSE)
```
