---
title: "Date extraction"
---

Language models are good at reading messy text and poor at calendar
arithmetic. A reliable extraction pipeline keeps those jobs apart: the model
*reads* the date parts out of a document, and R code assembles, validates and
compares the dates. This article adapts the pattern from the TypeSafe date
extraction cookbook. Chunks are not evaluated here.

```{r, eval = FALSE}
library(jevr)
```

## One Choice per date part

Ask for the year, month and day as three separate Choice questions. Each
Choice includes an explicit `not_stated` option, so a document that says
"sometime in March" can answer "March" for the month and "not stated" for the
day instead of being forced to invent one.

```{r, eval = FALSE}
months <- c(
  "01" = "January", "02" = "February", "03" = "March", "04" = "April",
  "05" = "May", "06" = "June", "07" = "July", "08" = "August",
  "09" = "September", "10" = "October", "11" = "November", "12" = "December"
)
years <- setNames(as.character(2023:2027), 2023:2027)
days  <- setNames(sprintf("Day %d of the month", 1:31), sprintf("%02d", 1:31))

with_not_stated <- function(options) {
  c(options, not_stated = "The document does not say")
}

questions <- list(
  year = jev_choice(
    "In what year does the invoice say payment is due?",
    with_not_stated(years)
  ),
  month = jev_choice(
    "In what month does the invoice say payment is due?",
    with_not_stated(months)
  ),
  day = jev_choice(
    "On what day of the month does the invoice say payment is due?",
    with_not_stated(days)
  )
)
```

A Choice may have up to 255 options, so the day question fits comfortably.
Keep the option names machine friendly (`"03"`, `"not_stated"`); the
descriptions are what the model reads.

## Extract from many documents

```{r, eval = FALSE}
documents <- c(
  inv_1 = "Payment is due by 14 March 2025. Thank you for your business.",
  inv_2 = "Please settle this invoice within thirty days. Ref 8812.",
  inv_3 = "Due: 2025-11-02"
)

results <- jev_map(documents, questions, concurrency = 3)
```

## Assemble and compare in code

The long format has one row per option; `selected` marks the chosen one. Pull
the chosen option per state and question, then build the date in R.

```{r, eval = FALSE}
long <- as.data.frame(results, format = "long")
chosen <- long[long$selected %in% TRUE, c("state_id", "question_id", "option")]

parts <- reshape(
  chosen,
  idvar = "state_id", timevar = "question_id", direction = "wide"
)
names(parts) <- sub("^option\\.", "", names(parts))

complete <- parts$year != "not_stated" & parts$month != "not_stated" &
  parts$day != "not_stated"
text <- paste(parts$year, parts$month, parts$day, sep = "-")
parts$due <- as.Date(ifelse(complete, text, NA_character_), format = "%Y-%m-%d")
parts
```

Passing `format` makes impossible combinations such as `2025-02-30` come
back as `NA` instead of raising an error. Treat every `NA` in `due` as "needs
review"; it means either that a part was not stated or that the parts do not
form a real date.

Now the arithmetic the model never does:

```{r, eval = FALSE}
parts$days_until_due <- as.numeric(parts$due - Sys.Date())
parts$overdue <- parts$due < Sys.Date()
```

## Gate review on confidence

Each Choice answer carries a confidence. Cut it into routing bands with
`jev_band()`. The package supplies no default thresholds; pick them from
labelled documents of your own.

```{r, eval = FALSE}
band_for <- function(id) {
  jev_band(
    results,
    question = id,
    breaks = c(0, 0.6, 0.85, 1),
    labels = c("review", "check", "auto")
  )
}

bands <- data.frame(
  state_id = names(results),
  year  = band_for("year"),
  month = band_for("month"),
  day   = band_for("day"),
  row.names = NULL
)
```

The result is a factor named by state ID (`NA` for failed states or answers
without a confidence). Combine the parts conservatively: a date is only as
trustworthy as its weakest part.

```{r, eval = FALSE}
rank_of <- function(x) match(as.character(x), c("review", "check", "auto"))
weakest <- pmin(rank_of(bands$year), rank_of(bands$month), rank_of(bands$day))
bands$route <- factor(c("review", "check", "auto")[weakest],
                      levels = c("review", "check", "auto"))

needs_review <- merge(parts, bands, by = "state_id")
needs_review <- needs_review[is.na(needs_review$due) |
                               needs_review$route %in% c("review", NA), ]
```

Missing dates (`NA`) and `NA` bands both go to review: an absent confidence is
not a high one.

## Reading the probabilities

When a document is ambiguous you can see it in the distribution rather than
only in the winner:

```{r, eval = FALSE}
subset(long, state_id == "inv_2" & question_id == "month" & probability > 0.05,
       select = c(option, probability, selected))
```

If `not_stated` carries most of the mass, the right behaviour is to leave the
date empty and route the document to a person, which the code above does.
