---
title: "Endpoints and gateways"
---

The TypeSafe System One wire format (`POST <base_url>/v1/systemone`) is served
by TypeSafe itself and by several AI gateways. jevr treats each server as an
*endpoint*: a base URL, the name of the environment variable that holds the
key, and a default model. None of the code below is evaluated when this site
is built, because every call needs a network connection and an API key.

```{r, eval = FALSE}
library(jevr)

questions <- list(
  urgent = jev_noul("Does this request convey urgency?")
)
state <- "My payouts have been failing for three days."
```

## Presets and environment variables

`provider =` accepts a preset name. Each preset reads its key from an
environment variable when the request is built; the key is never stored in
the endpoint or in results.

| Preset | Base URL | Key variable | Default model |
|---|---|---|---|
| `"typesafe"` | `https://api.typesafe.ai` | `TYPESAFE_API_KEY` | `jev-latest` |
| `"openrouter"` | `https://openrouter.ai/api` | `OPENROUTER_API_KEY` | `~typesafe/jev-latest` |
| `"vercel"` | `https://ai-gateway.vercel.sh/typesafe` | `AI_GATEWAY_API_KEY` | `typesafe-ai/jev` |
| `"pydantic"` | `https://gateway-us.pydantic.dev/proxy/typesafe` | `PYDANTIC_AI_GATEWAY_API_KEY` | `jev-latest` |

```{r, eval = FALSE}
Sys.setenv(OPENROUTER_API_KEY = "<your key>")  # or set it in ~/.Renviron

jev_ask(state, questions, provider = "openrouter")
```

The `"typesafe"` preset also honours `TYPESAFE_BASE_URL`, which is handy for
a staging server or a local proxy. The legacy preset
`"openrouter_decisions"` still reaches OpenRouter's alpha Decisions endpoint;
it is kept for compatibility, drops extra question fields, and is no longer
the way to use OpenRouter.

For any other TypeSafe-format server, describe it with `jev_endpoint()` and
pass the object as `provider`:

```{r, eval = FALSE}
gateway <- jev_endpoint(
  base_url = "https://gateway.example.com/typesafe",
  api_key_env = "EXAMPLE_GATEWAY_API_KEY",
  model = "jev-latest",
  headers = list("X-Team" = "support"),
  name = "example-gateway"
)
gateway

jev_ask(state, questions, provider = gateway)
```

`base_url` must not include `/v1/systemone`, and `Authorization` cannot be
set through `headers`; the bearer token always comes from `api_key_env`.

## Listing models

`jev_models()` calls `GET <base_url>/v1/models` and returns a data frame with
`name`, `description` and `release_date`.

```{r, eval = FALSE}
jev_models()
jev_models("vercel")
```

OpenRouter publishes its catalogue in a different shape, so asking for the
OpenRouter presets raises a `jev_unsupported_error` before any request is
sent. Browse <https://openrouter.ai/typesafe> instead.

## Cost and routing metadata

Gateways report what a request cost and where it was routed. jevr keeps both
without interpreting them beyond a single normalized number.

```{r, eval = FALSE}
response <- jev_ask(state, questions, provider = "vercel")

response$usage$cost                  # USD, normalized; absent if unreported
response$metadata$provider_metadata  # gateway metadata, verbatim
response$metadata$final_provider     # who actually served the request
response$metadata$generation_id      # gateway id for support and audits
```

`usage$cost` comes from `usage.cost` (OpenRouter) or from the gateway's
`provider_metadata` (Vercel, converted from a string to a number). For
`jev_map()` the per-request ledger holds the same number, with `NA` where it
was not reported, and `summary()` adds them up.

```{r, eval = FALSE}
results <- jev_map(
  c(a = state, b = "Refund still pending."),
  questions,
  provider = "vercel"
)
summary(results)$observed_cost
attr(results, "requests")[, c("state_id", "attempt", "cost")]
```

## Which model answered?

`jev-latest` is a moving alias. Every response records both sides:

```{r, eval = FALSE}
response$metadata$requested_model  # what you asked for, e.g. "jev-latest"
response$model                     # versioned model the server reported

as.data.frame(results)[, c("state_id", "requested_model", "answered_model",
                           "is_alias")]
```

`is_alias` is `TRUE` for moving names (`jev-latest`, `jev-preview`, OpenRouter
`~` aliases, the unversioned Vercel `typesafe-ai/jev`). Execution ids are
computed before sending, so they use the *requested* model only. Two runs
with `jev-latest` therefore share ids even if different versions answered;
do not reuse alias-based results across sessions as if they were identical.

If more than one version answers inside a single `jev_map()` run, jevr warns
with class `jev_model_drift_warning`, and `summary(results)$answered_models`
lists everything it saw. For reproducible work, pin a versioned ID taken from
a previous response or from `jev_models()`:

```{r, eval = FALSE}
pinned <- response$model   # a specific version string reported by the server
results <- jev_map(c(a = state, b = "Refund still pending."), questions,
                   model = pinned)

withCallingHandlers(
  jev_map(c(a = state, b = "Refund still pending."), questions),
  jev_model_drift_warning = function(w) {
    message("Drift: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
```

## Decision fallbacks on Vercel

Vercel AI Gateway can rerun an uncertain decision with a stronger model.
`jev_fallback()` describes the trigger and the second model; pass it as
`fallback =`. Only endpoints with the `"fallback"` capability accept it (the
`"vercel"` preset, or a custom endpoint created with
`capabilities = "fallback"`); otherwise jevr raises a `jev_input_error`
before sending anything.

```{r, eval = FALSE}
questions <- list(
  intent = jev_choice(
    "Which team should handle this message?",
    c(billing = "Charges and refunds", account = "Sign-in and access")
  ),
  refund = jev_noul("Is the customer asking for money back?")
)

fb <- jev_fallback("openai/gpt-6-astra", "intent", confidence_below = 0.6)

response <- jev_ask(
  "I was charged twice and cannot sign in.",
  questions,
  provider = "vercel",
  fallback = fb
)
```

Conditions can be combined with `jev_when()` and the compound helpers
`jev_when_any()`, `jev_when_all()` and `jev_when_at_least()`. Noul questions
use `probability_between`, an inclusive band around 0.5:

```{r, eval = FALSE}
fb <- jev_fallback(
  "openai/gpt-6-astra",
  when = jev_when_any(
    jev_when("intent", confidence_below = 0.6),
    jev_when("refund", probability_between = c(0.4, 0.6))
  ),
  on_error = "anthropic/claude-sonnet-5"
)
```

When the fallback runs, the gateway reruns the state with *all* questions and
returns that complete result; answers are not merged, and both stages are
billed (`usage$cost` is already the total). jevr parses the gateway's
decision-fallback headers into `response$metadata$fallback`:

```{r, eval = FALSE}
fallback <- response$metadata$fallback  # absent when it did not trigger
fallback$triggered
fallback$primary_model
fallback$final_model
fallback$triggering_questions
response$model                          # names the fallback model if it answered
```

A language-model fallback has no probability distribution to report. The
gateway signals this with `confidence: 0` and empty probabilities, which jevr
turns into `NA` confidence and `NA` probabilities (with
`confidence_unavailable = TRUE`) rather than a misleading zero. Handle it
explicitly before routing on confidence:

```{r, eval = FALSE}
answer <- response$answers$intent
if (isTRUE(answer$confidence_unavailable)) {
  # the final answer came from a fallback LLM: choice is usable, confidence is not
  answer$choice
}

# NA when confidence is unavailable
jev_band(response, "intent", breaks = c(0, 0.6, 1), labels = c("review", "auto"))
```

Noul answers are unaffected by the sentinel.
