---
title: "Reranking"
---

A reranker orders candidates by how well they answer a query. With System One
the model does only the part that needs judgment: *how relevant is this one
candidate to this one query?* Retrieval, tie-breaking, thresholds and the
final sort stay in ordinary R code. This article adapts the idea from the
TypeSafe reranking cookbook; chunks are not evaluated here.

```{r, eval = FALSE}
library(jevr)
```

## Retrieve in code, judge pairs with the model

Assume a first-stage retriever (BM25, vector search, a SQL filter) already
returned a handful of candidates per query. The model never sees the whole
list, so candidate order cannot bias it and the list can be as long as you
like.

```{r, eval = FALSE}
candidates <- data.frame(
  query_id = c("q1", "q1", "q1", "q2", "q2"),
  doc_id   = c("d1", "d2", "d3", "d4", "d5"),
  query    = c(rep("how do I reset my password", 3),
               rep("export invoices to csv", 2)),
  text     = c(
    "Open Settings, choose Security, and select Reset password.",
    "Our pricing plans include a free tier.",
    "Passwords must contain at least twelve characters.",
    "Invoices can be downloaded as CSV from the Billing page.",
    "Contact sales for volume discounts."
  ),
  stringsAsFactors = FALSE
)
candidates$key <- paste(candidates$query_id, candidates$doc_id, sep = "/")
```

## One comparable question per pair

Each query-candidate pair becomes one *state*. Use a single Score question
with the same levels for every pair so the numbers are comparable across
states. Describe the levels concretely; an unanchored "how relevant" scale is
vague.

```{r, eval = FALSE}
questions <- list(
  relevance = jev_score(
    "How well does the document answer the query?",
    c(
      "Unrelated to the query",
      "Same topic but does not answer it",
      "Partially answers the query",
      "Directly and fully answers the query"
    )
  )
)

states <- lapply(seq_len(nrow(candidates)), function(i) {
  list(query = candidates$query[i], document = candidates$text[i])
})
names(states) <- candidates$key
```

A Noul ("Does this document answer the query?") works too when you only need
a filter. Score returns a probability-weighted position, so it separates
close candidates better.

## Run the pairs with `jev_map()`

```{r, eval = FALSE}
results <- jev_map(states, questions, concurrency = 4)
results
```

`jev_map()` keeps input order and the state IDs you named. A failed pair
becomes a failed result rather than aborting the run.

## Rank in code

The long format returns one row per state, question and level. For a Score
question the `score` column holds the probability-weighted expectation, so
keep one row per state before joining.

```{r, eval = FALSE}
long <- as.data.frame(results, format = "long")
rel <- long[long$question_id == "relevance" & !duplicated(long$state_id), ]

scores <- data.frame(key = rel$state_id, score = rel$score, model = rel$model)
ranked <- merge(candidates, scores, by = "key")

ranked <- ranked[order(ranked$query_id, -ranked$score), ]
ranked$rank <- ave(-ranked$score, ranked$query_id, FUN = rank)
ranked[, c("query_id", "doc_id", "score", "rank")]
```

Everything after `jev_map()` is plain data-frame work, so you can add your
own rules: break ties by recency, keep the top `k`, or drop documents scoring
under a threshold you chose by looking at labelled examples.

```{r, eval = FALSE}
top <- do.call(rbind, lapply(split(ranked, ranked$query_id), head, n = 3))
```

## Failed pairs and provenance

Check status before trusting a ranking. In long format a failed pair has a
single row with `NA` score and the error in `error_class`.

```{r, eval = FALSE}
subset(long, status != "success",
       select = c(state_id, status, error_class, error_message))

unique(long$model)  # versioned model(s) that produced the scores
summary(results)$answered_models
```

Scores from different model versions are not strictly comparable. If a run
reports more than one answered model (a `jev_model_drift_warning`), re-run it
pinned to one versioned model, as described in the endpoints article.

## Why not rank one list at once?

Giving the model every candidate in one prompt couples each answer to
position and to the other candidates. Independent pair scores are
cacheable (execution ids are stable per state and spec), parallel, and
auditable one row at a time.
